---
title: "qc droplets"
author: "Bjorn"
date: "10/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import the raw Cell Ranger matrix for 1,000 cells into R; use DropletUtils.
```{r}
library(DropletUtils)
sce <- read10xCounts("data/pbmc_1k_v3_raw", sample.names = "PBMC")
sce

View(as.data.frame(colData(sce)))
```

Remove droplets without any count at all.
```{r}
#we first sum for each column (=cells) to see which cell has 0 coints
sums <- colSums(assay(sce, 'counts'))
#we then replace the original object with the one that has counts
sce <- sce[, sums > 0]
sce
```
Run DropletUtils::emptyDrops() .
```{r}
assay(sce)
empty_drops <- DropletUtils::emptyDrops(m = assay(sce, 'counts'))
empty_drops

library(ggplot2)
plot1 <- ggplot(as.data.frame(empty_drops), aes(x = Total)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_log10() +
 geom_vline(xintercept = 100)
plot1

#we can calculate the number of cells with a Total of more than 100
sum(empty_drops$Total > 100)
#this detects 1365 cells

#another method using the FDR
is.cell <- empty_drops$FDR <= 0.01
sum(is.cell, na.rm=TRUE)
#it detects 1205 cells at a FDR of <=0.01
#as this is a machine learning algorithm, we get slightly different results

#if we wanted to generate an elbow plot, we need this
#the rank function ranks cells from the highest to the lowest total count and adds it to empty_drops (in a new rank column)
empty_drops$Rank <- rank(-empty_drops$Total)
empty_drops

#we can now plot the rank on the x-axis and the Total on the y-axis
plot2 <- ggplot(as.data.frame(empty_drops), aes(x = Rank, y = Total)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
 geom_hline(yintercept = 100)
plot2
#we can use 100 as a cutoff
```
Remove empty droplets from the data sets.
```{r}
#We can use which to exclude NA and FALSE values; TRUE is when FDR <= 0.01
which(is.cell)

#we now want to remove empty droplets from the data set using subsetting for the columns
sce <- sce[, which(is.cell)]
sce
#this sce object now only contains 1205 cells
```
Doublets. Continue using the data set after excluding empty droplets. Run scDblFinder.::scDblFinder().
```{r}
library(scDblFinder)
sce <- scDblFinder::scDblFinder(sce)
#it tells us that (for some reason) 5000 artificial doublets were used, 15 doublets were called
#we can have a look at the new object:
table(rowData(sce)$scDblFinder.selected)
#this tells us that 32538 are FALSE, 1000 TRUE (this is the default of scDblFinder)
colData(sce)$scDblFinder.originAmbiguous
colData(sce)$scDblFinder.class
```
One version to plot is plotDoubletMap
```{r}
plotDoubletMap(sce)
```
Remove doublets from the data sets.
```{r}
#we create a new thing called is.doublet, subsetting it based on the scDblFinder.Class column, which says "singlet" for singlets and "doublets" for doublets 
is.doublet <- colData(sce)$scDblFinder.class
table(is.doublet)

#we are then removing the doublets from the object
sce <- sce[, is.doublet == "singlet"]
sce
#the object retains everything that doesn't contain 
```
Ambient RNA. Run celda::decontX().
```{r}
library(celda)
sce <- decontX(sce)
colData(sce)

#we can visualise the percentage of contamination
plotDecontXContamination(sce)

table(colData(sce)$decontX_clusters)
```

